---
title: "Assignment 2 (Exercise 1)"
author: "201843064"
date: "October 31st 2023"
output: html_document
---

```{r setup, include=FALSE} 
library(tidyr)
library(tidyverse)
library(purrr)
library(dplyr)

knitr::opts_chunk$set(echo = FALSE) 
```

## Function pop.change()
# Calculates population change and inserts outcome into text chunk
This function was built to work with the 'population' dataset in tidry. It takes a dataset that includes population data for different countries across different years and calculates average population change between the first and the last year for which data is available. It then inserts the calculated result in a text chunk. This function may for instance be useful for a yearly report that takes up-to-date population figures and calculates population change against a baseline.

The example below shows the output of the function for a country with below average population growth (Germany) and a country with above average population growth (Rwanda).

```{r plot_example, echo=FALSE}
# Load dataset 'population' from tidyr
data('population')

# Create function which calculates population change between two years and prints text
 # Input: population dataset and country of interest
 # Output: calculated percentage change in population and country rank inside text snippet

# !Warning!
  # This function was written using column names from the 'population' dataset
  # found in the 'tidyr' package. In order to use this function with a different dataset some
  # columns may need to be renamed. The code below shows how to rename columns to match the
  # names used in the function below
  # names(data)[names(data) == 'old.country.var.name'] <- 'country'
  # names(data)[names(data) == 'old.year.var.name'] <- 'year'
  # names(data)[names(data) == 'old.year.var.name'] <- 'population'

pop.change <- function(data, country) {

  ## Preparation
  # Define helper function
  # Code for helper function was taken from (https://stackoverflow.com/questions/5831794/opposite-of-in-exclude-rows-with-values-specified-in-a-vector)
  
  `%not_in%` <- purrr::negate(`%in%`)
  
  ## Check Assumption
  # Assumption 1: population data is available for all countries across the years of interest.
  # In this case, we are interested in the first and last year for which data is available
  # in the dataset.

  # define first and last year 
  firstyear <- min(data$year)
  lastyear <- max(data$year)
  
  # check if data is missing for years of interest
  check.df <- data %>%
    arrange(country, year) %>% 
    group_by(country) %>%
    summarise(check.year = if_else((firstyear %in% year && lastyear %in% year), 'Complete','Missing'),
              check.data = if_else(is.na(population[year == first(year)])|
                                          is.na(population[year == last(year)]), 'Missing','Complete' ))
  
  no.data <- check.df %>% filter(check.year == "Missing" | check.data == "Missing") 
  
  ## removes countries for which data is missing 
  data <- subset(data, data$country %not_in% no.data$country)
  
 ## Calculation   
 # calculate population change in percent
change.df <- data  %>%
  arrange(country, year) %>% # arrange ensures that the order is correct (earlier to later years)
  group_by(country) %>%
  summarise(perct_change = (last(population) - first(population))/first(population)*100) %>%
  ungroup()

 # calculate mean percentage change across all countries 
mean.change <- mean(change.df$perct_change)

 # calculate percentage change for country defined
country.change <- change.df$perct_change[change.df$country == country]

 # calculate rank of each country 
change.df$low2high <- rank(change.df$perct_change)
change.df$high2low <- rank(- change.df$perct_change)


##Output

if (country.change > mean.change) {
  
 # calculate rank of country defined high to low
country.rank <- change.df$high2low[change.df$country == country]

 # print text
print(paste0("On average population size changed by ", round(mean(change.df$perct_change), 2),
             " percent between ", firstyear," and ", lastyear, " across the countries in the dataset.", " At ", round(country.change,2)," percent population change ", country ," was ", round((country.change - mean.change),2) ," percent above the average. This places ", country," as the country with the ", country.rank, ". largest increase in population numbers."))
  
print(ifelse(length(no.data)>0,
             paste0("Please note that the following countries were removed due to missing data: ",
                    paste(no.data$country, collapse = ", "), "."),
             paste0("")))

} else {

 # calculate rank of country defined low to high
country.rank <- change.df$low2high[change.df$country == country]

print(paste0("On average population size changed by ", round(mean(change.df$perct_change), 2),
             " percent between ", firstyear," and ", lastyear, " across the countries in the dataset.", " At ", round(country.change,2)," percent population change ", country ," was ", round((country.change - mean.change),2) ," percent below the average. This places ", country," as the country with the ", country.rank, ". lowest increase in population numbers."))

print(ifelse(length(no.data)>0,
             paste0("Please note that the following countries were removed due to missing data: ",
                    paste(no.data$country, collapse = ", "), "."),
             paste0("")))

 }
}

# run function for country with below average population growth = Germany 
pop.change(population, "Germany")


# run function for country with above average population growth = Rwanda
pop.change(population, "Rwanda")


```
```
## Appendix: All code in this assignment

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE} 
# this chunk generates the complete code appendix. 
# eval=FALSE tells R not to run (``evaluate'') the code here (it was already run before).
```
